---
title: Data Cleaning, Data Exploration, Data Modelling and Model Deployment using
  R
author: "Famous Salami | salamifamous@gmail.com"
date: "15-10-2023"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

## Dataset: Records from an insurance company's recently insured drivers.
### .
### Importing Libraries
```{r include = F} 
getwd()
library(caret)
library(ggplot2)
library(stringr)
library(dplyr)
library(corrplot)
library(Hmisc)
library(formattable)
```

```{r}
#clears the R environment
rm(list=ls())
```

```{r}
#sets working directory to current directory
setwd(getwd())
```

### .
### Loading the dataset

```{r}
#Reading the data file
insuredb <- read.csv("insurance.csv", stringsAsFactors = T)
```

### .
### Inspecting and understanding the data

```{r}
#structure of the data
str(insuredb)
```


```{r}
#in-depth view of the data
describe(insuredb)
```

```{r}
#summary of the data, for insight in slightly different form
summary(insuredb)
```
### .
#### Observing the first 6 rows
```{r}
# Observing the first 6 rows
head(insuredb)
```
```{r}
min(insuredb$driver_age)
```

The dataset comprises 1,000 observations across 13 variables (columns). It comprises of information on insurance premiums, driver demographics, accident rates and car-related details within the United Kingdom.

The insurance premiums range from £150.32 to £780.42, showcasing variability in policy costs. Driver ages span within the range of 17 to 86 years, with the average age being around 47.

'crime_area' contains 22 missing values.

'annual_mileage' also contains 32 missing values identified as NAs

The 'gender' variable is meant to comprise categorical data but is messed up with some inconsistent values.

More insights would be shared on this data with visualization tool, Power BI.


### .
### Data Cleaning and Preparation

The features 'car_reg' and 'country' would not be suitable as predictor because the columns contain 1,000 unique values which is same as total observations and cannot guarantee suitable scientific prediction.

```{r}
insuredb$car_reg <- NULL #to remove the car_reg column

insuredb$country <- NULL #to remove the country column

summary(insuredb)
```
#### .
##### Features 'annual_mileage' and 'crime_area' has NULL values, needs to be fixed

```{r}
#To determine best value to fix for the missing values of 'annual_mileage'
ggplot(data = insuredb, aes(x = annual_mileage)) + 
  geom_histogram(bins = 30, na.rm = TRUE, color="black", fill="white", position = "stack")

```

The histogram above is normally symmetric, therefore the mean of the column values is suitable as a replacement for the missing values.

```{r}
#get the mean of the values
getMean = mean(insuredb$annual_mileage, na.rm = TRUE)

insuredb$annual_mileage[is.na(insuredb$annual_mileage)] <- getMean
#summary(insuredb)

```

```{r}
#To determine best value to fix for the missing values of 'crime_area'

ggplot(data = insuredb, aes(x = crime_area, y = insurance)) + geom_boxplot()

```
 
```{r}
#to replace missing values with 'low' being the median
insuredb$crime_area[(insuredb$crime_area == '')] <- 'low'

insuredb$crime_area = droplevels(insuredb$crime_area)

```

#### .
#### Transforming the 'gender' column to have consistent values, 'Male' and 'Female'
```{r}
#gender column transformation
prepGender <- function(genderStr){
  cSgenderStrtr <- as.character(genderStr)
  if(str_length(genderStr) == 1){
    if(genderStr == 'M') genderStr = as.factor('Male')
    if(genderStr == 'F') genderStr = as.factor('Female')
  }else{
    genderStr = str_to_sentence(genderStr)
  }
  genderStr = as.factor(genderStr)
}

insuredb$gender = sapply(insuredb$gender, prepGender)
#summary(insuredb$gender)
```

#### .
#### Transforming the 'car_value' column to have consistent numerical data

```{r}
#car_value column transformation to get rid of £ and K from the values
carValue <- function(cStr){
  cStr <- as.character(cStr)
  if(str_detect(cStr, 'K')){
    newVal = str_remove_all(cStr, "[£K]")
    newVal = as.numeric(newVal)
    cStr = newVal * 1000
  }
  if(str_detect(cStr, ",")){
    newVal = str_remove_all(cStr, "[£,]")
    cStr = newVal
  }  
  cStr = as.numeric(cStr)
}

insuredb$car_value = sapply(insuredb$car_value, carValue)
#summary(insuredb)
```

#### .
#### Transforming the 'car_age' column to have consistent numerical data

```{r}
#car_age column transformation to get rid of Y(ear) and M(onth)

carAge <- function(cStr){
  cStr <- as.character(cStr)
  if(str_detect(cStr, ' ')){
    spVal = unlist(str_split(cStr, " "))
    yr = str_remove_all(spVal[1], 'Y')
    mth = str_remove_all(spVal[2], 'M')
    nyr = as.numeric(yr)
    nmth = as.numeric(mth)
    yr = 12*nyr
    cStr = yr + nmth
  }
  cStr = as.numeric(cStr)
}

insuredb$car_age = sapply(insuredb$car_age, carAge)

```

### .
### Exploratory Data Analysis

```{r}

ggplot(data = insuredb, aes(x = insurance)) + 
  geom_histogram(bins = 32, na.rm = TRUE,color="black", fill="white")
```

The histogram of insurance is skewed(positive) to the right.

Most of the drivers spent less than the average insurance premium of (£310.8) while a few of the drivers spent more, up to (£700+).

```{r}
ggplot(data = insuredb, aes(x = annual_mileage, color=gender)) + geom_histogram(bins = 30, na.rm = TRUE, fill="white")

```

The histogram of annual_mileage is symmetrical. Shows a perfect balance in the miles traveled by drivers annually.

• Key takeaway: for every person who drives more miles than the average (10,055 miles) annually, there's another person who drives fewer miles than the average.


#### .
#### Determining gender with highest annual mileage

```{r}
# Calculate the gender with the highest number of accidents
result <- aggregate(annual_mileage ~ gender, data = insuredb, FUN = sum)
#result
male_miles <- result[result$gender == "Male", "annual_mileage"]
female_miles <- result[result$gender == "Female", "annual_mileage"]

cat("Total number miles traveled by males annually is:", 
    sprintf("%s", prettyNum(male_miles, big.mark = ",", decimal.mark = ".")), ' miles\n')
cat("Total number miles traveled by females annually is:", 
    sprintf("%s", prettyNum(female_miles, big.mark = ",", decimal.mark = ".")),'miles')

ggplot(result, aes(x = gender, y = annual_mileage)) + geom_bar(stat = "identity", fill = "orange") +
  labs(x = "Gender", y = "Annual Mileage")
```

#### .
#### Determining gender with highest number of accidents

```{r}
# Calculate the gender with the highest number of accidents
result <- aggregate(num_accident ~ gender, data = insuredb, FUN = sum)
#result
male_accidents <- result[result$gender == "Male", "num_accident"]
female_accidents <- result[result$gender == "Female", "num_accident"]

cat("Total number accidents by males is:", male_accidents, '\n')
cat("Total number accidents by females is:", female_accidents)

ggplot(result, aes(x = gender, y = num_accident)) + geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Gender", y = "Number of accidents")
```

#### .
#### Determining gender with highest cars value (most expensive cars)

```{r}
# Calculate the gender with the highest number of accidents
result <- aggregate(car_value ~ gender, data = insuredb, FUN = sum)
#result
male_cars <- result[result$gender == "Male", "car_value"]
female_cars <- result[result$gender == "Female", "car_value"]

cat("Total males cars value is:", sprintf("£%s", prettyNum(male_cars, big.mark = ",")), '\n')
cat("Total females cars value is:", sprintf("£%s", prettyNum(female_cars, big.mark = ",")))

ggplot(result, aes(x = gender, y = sprintf("£%s", prettyNum(car_value, big.mark = ",")))) +
  geom_bar(stat = "identity", fill = "magenta") +
  labs(x = "Gender", y = "Number of accidents")
```

#### .
#### Determining insurance group with highest aggregated insurance premiums

```{r}
#result <- insuredb[which.max(insuredb$num_accident), "gender"]
result <- aggregate(insurance ~ insurance_group, data = insuredb, FUN = sum)
#result
group1_insurance <- result[result$insurance_group == "Group 1", "insurance"]
group2_insurance <- result[result$insurance_group == "Group 2", "insurance"]
group3_insurance <- result[result$insurance_group == "Group 3", "insurance"]

g1_2Currency <- sprintf("£%s", prettyNum(group1_insurance, big.mark = ",", decimal.mark = "."))
g2_2Currency <- sprintf("£%s", prettyNum(group2_insurance, big.mark = ",", decimal.mark = "."))
g3_2Currency <- sprintf("£%s", prettyNum(group3_insurance, big.mark = ",", decimal.mark = "."))

cat("Total insurance premium paid by Group 1 is:", g1_2Currency, '\n')
cat("Total insurance premium paid by Group 2 is:", g2_2Currency, '\n')
cat("Total insurance premium paid by Group 3 is:", g3_2Currency)

ggplot(result, aes(x = insurance_group, y = insurance)) + geom_bar(stat = "identity", fill = "purple") +
labs(x = "Insurance group", y = "Insurance")


```

#### .
#### Determining relationship between number of accident and driver's annual mileage

```{r}
ggplot(insuredb, aes(x = annual_mileage, y = num_accident)) +
  geom_point() +
  labs(x = "Annual Mileage", y = "Number of Accidents") +
  ggtitle("Number of Accidents vs. Annual Mileage")

```

The plot reveals that as the number of miles driven annually increases, the number of accidents also increases. Therefore, if a driver drives a lot of miles each year, there's a higher chance he/she may be involved in more accidents and vice versa.

• Application: An important insight for insurance or safety considerations. 

• Decision support: It is important to consider annual mileage when assessing the risk of accidents in the context of insurance or safety planning.

#### .
#### Write the final transformed dataset to current directory for visualization in Power BI

```{r}

write.csv(insuredb, file.path(getwd(), "new_insurancedb.csv"), row.names = FALSE)

```

#### .
##### Correlation between insurance and other key variables

```{r}
#corrplot(cor(insuredb[(1:7)]))

corrplot(cor(insuredb[(1:7)]), method = "number")
```

### .
### Conclusion
• Insurance vs. Driver Age: Negative correlation between insurance costs and driver age. As a driver's age increases, insurance costs tend to decrease.

• Insurance vs. Car Value: Positive correlation between insurance costs and the value of the car. Cars with higher value have higher insurance costs.

• Insurance vs. Number of Accidents: Positive correlation between insurance costs and the number of accidents. For drivers involved in more accidents, their insurance costs seem to be higher.

• Insurance vs. Annual Mileage: Positive correlation between insurance costs and the number of miles driven annually. The more miles driven, the higher the insurance costs.

• Insurance vs. Car Age: Weak negative correlation between insurance costs and the age of the car. As the car gets older, insurance costs may slightly decrease.

• The male gender are significantly involved in more accidents than the females, this might be because the males travel more miles than the females annually.

• The male gender drives more expensive cars than the females but with a moderate margin.



### .
### Future Works
My next work on this would be to:

• build a predictive model

• build a risk assessment model

• create a Power BI visualization









